\bf Visualizing Dynamic Social Network Models

### 2018-11-12 paper-child.Rnw
starting to generalize

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\textwidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, dev="cairo_pdf", fig.width = 6, fig.height = 6)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
@

@@

<<getdata , echo=FALSE>>=
set.seed(823746)
load("data/ansnullpaper.rda")
ansnullchains <- get_chain_info(ansnull)
load("data/M1sims1000.RData")
load("data/M2sims1000.RData")
load("data/M3sims1000.RData")
@

\section{Introduction \& Background}
% peices of old intro
%Social networks, such as collaboration networks between academic researchers or friendship networks 
% new intro 
%As humans and researchers, it is our instint to try to learn as much as we can about the world around us. 
As members of the human family, one of the hardest things we try to do is understand our relationships with one another, whether as friends, family, or colleagues. These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals. There are also statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}. These static models, however, are not optimal for modeling ``real-world" social networks because they do not account for changes in the network as time passes. Models for \textit{dynamic social networks} are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time. %., and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms. %Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our relationships with each other.

\par One set of models for dynamic social networks are stochastic actor-oriented models (SAOMs), introduced in \citet{saompaper}. These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree, common in ERGMs.
%These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information. Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example 
With these models, we can examine assumptions made about social networks, i.e. that people with common interests are more likely to become friends. For example, we may assume that teenagers who drink heavily are more likely to be friends with other tennagers who drink heavily. The SAOM family allows us to incorporate this behavioral information on shared interests into the modeling process.

@@

<<toyex, echo = FALSE>>=
set.seed(81393)
nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
@
<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the corresponding adjacency matrix visualization.", out.width='.49\\textwidth', fig.height=4>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) +
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
  ThemeNet +
  theme(plot.title = element_text(size = 20)) +
  labs(title = "Node-Link Diagram")
#ggsave("figure/toynetnld.pdf",plot = nld, width = 4, height = 4)

ggplot(data = toynet$edges) +
  #geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  geom_tile(aes(x = to, y = -from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) +
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) +
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  ThemeNoNet +
  theme(panel.grid = element_blank(), aspect.ratio = 1,
        plot.title = element_text(size = 20),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        axis.text.x  = element_text(size = 18),
        axis.text.y  = element_text(size = 18)) +
  labs(title = "Adjacency Matrix Visualization", x = "\nto (alter)", y = "from (ego)\n")
@
%\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. These probabilities are defined via the objective function. Let $x(i \leadsto j)$ be the network identical to network $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$; i.e.\ the presence or absence of node $x_{ij}$ is flipped in the new network. The probability that the tie $x_{ij}$ changes is defined as:
\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
\end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
%When the value of the objective function is high at the current state, the probability of not making a change is also high.
We explore these probabilities further in Section~\ref{sec:algorithms}. The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change. For more details on CTMCs, see e.g. \citet{ctmc}. 
% \subsubsection{Continuous Time Markov Chain (CTMC)}
% \st{
% In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states for a directed network with $n$ nodes. We denote the state space as $\mathcal{X}$, a set which contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and there are $n(n-1)$ edge relationships because the network is directed and we exclude self-ties.  The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$. Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged. All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.
% 
% The two pieces of a SAOM, the rate function/parameter and the objective function, each contribute to the entries of the intensity matrix to describe the rate of change between two network states. The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.  Then the $bc^{th}$ entry of $Q$ is:
% \[ q_{bc} = \begin{cases}
%      q_{ij}= \alpha_m p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \{1, \dots, n\}\} \\
%       0 & \text{if } \sum_i \sum_j |x_{ij}^c -x_{ij}^b | > 1 \\
%       -\sum_{i\neq j} q_{ij} & \text{if } x^b = x^c
%    \end{cases}
% \]
% 
% Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$. This matrix $\mathbf{Q}$ is the foundation for estimation of a SAOM. %Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep{RSiena}, we have an accessible way to fit SAOMs to network observations.
% 
% %XXX HH can you help? I need a way better transition than that XXX
% }

%### A SAOM as a CTMC {#saomctmc}

%In order to fit this model definition back into the original context of the CTMC described in Section \@ref(dynamicnets), it must be written in terms of its intensity matrix, $\mathbf{Q}$.  This matrix describes the rate of change between states of the process. For networks, there are a very large number of possible states, $2^{n(n-1)}$, so the intensity matrix is a square matrix of that dimension. But, thanks to the property of SAOMs that the states are allowed to change only one tie at a time, there are only $n$ possible states given the current state, $n-1$ of which are uniquely determined by the node $i$ that is given the opportunity to change. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these represent the possible states that are one edge different from a given state, and the additional non-zero entry is for the state to remain the same. All other entries in a row are zero because those column states cannot be reached from the row state by just one change as dictated by the SAOM. The entries of $\mathbf{Q}$ are defined as follows: let $b \neq c \in \{1, 2, \dots, 2^{n(n-1)} \}$ be indices of two different possible states of the network, $x^b, x^c \in \mathcal{X}$.  Then the $bc^{th}$ entry of $Q$ is:
% \[ q(x^b, x^c) = \begin{cases}
%       q_{ij}(\alpha, \rho, \boldsymbol{\beta}, x^b) = \lambda_i(\alpha, \rho, x^b, m)p_{ij}(\boldsymbol{\beta}, x^b) & \text{if } x^c \in \{x^b(i \leadsto j) | \text{ any } i \neq j \in \mathcal{N}\} \\
%       0 & \text{if } x^c \text{ differs from } x^b \text{ by more than 1 tie} \\
%       -\sum_{i\neq j} q_{ij}(\alpha, \rho, \boldsymbol{\beta}, x^b)  & \text{if } x^b = x^c
%    \end{cases}
% \]
%
% Thus, the rate of change between any two states that differ by only one tie, $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.^[Just to be clear, the change is from $x^b_{ij}$ to $x^c_{ij} = 1 - x^b_{ij}$.] Furthermore, the theory of continuous time Markov chains gives that the matrix of transition probabilities between observation times $t_{m-1}$ and $t_{m}$ is dependent only on the difference between timepoints, $t_m - t_{m-1}$. Following the same definition for transition probabilities in Section \@ref(dynamicnets), the matrix of transition probabilities is
% $$e^{(t_m - t_{m-1})\mathbf{Q}},$$
% where $\mathbf{Q}$ is the matrix defined above and $e^X$ for a real or complex square matrix $X$ is equal to $\sum_{k=0}^{\infty} \frac{1}{k!} X^k$.

@@

%\subsection{Model Goodness-of-Fit}\label{sec:modgof}

%The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}. This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}. Examples of auxiliary statistics include the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine. The goodness-of-fit is evaluated as follows:
% \begin{enumerate}
% \item The auxiliary statistics are computed on the observed data and $N$ simulated observations from the fitted model. Typically, $N=1000$.
% \item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
% \item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2.
% \item The Mahalanobis distance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
% \item An empirical $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data. A SAOM is thus considered a good fit if $p$ is large.
% \end{enumerate}

% <<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network. \\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
% load("data/ansnullpaper.rda")
% library(lattice)
% gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
% gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
%
% sims1 <- data.frame(gof1$`Period 1`$Simulations)
% dat1 <- gof1$`Period 1`$Observations
% dat1 %>% data.frame() %>% gather(outdegree, val) %>%
%   mutate(outdegree = parse_number(outdegree)-1) -> dat1
% p1 <- sims1 %>%
%   mutate(sim = row_number()) %>%
%   gather(outdegree, val, X1:X9) %>%
%   mutate(outdegree = parse_number(outdegree) - 1) %>%
%   filter(outdegree <= 6) %>%
%   ggplot(aes(x = outdegree, y = val)) +
%   geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) +
%   geom_violin(aes(group = outdegree), bw = 2, fill = NA) +
%   geom_point(data = dat1, color = 'red') +
%   geom_line(data= dat1, color = 'red') +
%   geom_text(data = dat1, aes(label = val), hjust = -.5) +
%   labs(x = "Outdegree (p = 0.154)", y = "Statistic",
%        title = "Goodness-of-Fit: Outdegree distribution period 1") +
%   ThemeNoNet
%
% sims2 <- data.frame(gof2$`Period 1`$Simulations)
% dat2 <- gof2$`Period 1`$Observations
% dat2 %>% data.frame() %>% gather(indegree, val) %>%
%   mutate(indegree = parse_number(indegree)-1) -> dat2
% p2 <- sims2 %>%
%   mutate(sim = row_number()) %>%
%   gather(indegree, val, X1:X9) %>%
%   mutate(indegree = parse_number(indegree) - 1) %>%
%   filter(indegree <= 6) %>%
%   ggplot(aes(x = indegree, y = val)) +
%   geom_boxplot(aes(group = indegree), outlier.shape = "x", outlier.size = 5) +
%   geom_violin(aes(group = indegree), bw = 2, fill = NA) +
%   geom_point(data = dat2, color = 'red') +
%   geom_line(data= dat2, color = 'red') +
%   geom_text(data = dat2, aes(label = val), hjust = -.5) +
%   labs(x = "Indegree (p = 0.063)", y = "Statistic",
%        title = "Goodness-of-Fit: Indegree distribution period 1") +
%   ThemeNoNet
%
%
% gridExtra::grid.arrange(p1, p2)
% # sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
% # [1] 0.154
% # > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
% # [1] 0.063
%
% @

%The \texttt{plot.sienaGOF()} function allows us to visualize this fit. %In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model.
%This function draws a box plot and a violin plot at each value of the statistic of interest observed in the simulations: \st{on the $x$-axis, the out(in)degrees observed in the data ($0, 1, 2, 3, \dots$), and on the $y$-axis, the cumulative number of times that out(in)degree value appears in the simulations and the data.} In order to compare the distribution of the counts of nodes with the specified degree as calculated on the simulated networks to the counts observed in the true data, red points connected by red lines representing the observed values are superimposed on the boxplots. If the red points lie "well within" the simulated values, the model is a good fit to the data. \st{This plot is not shown because it is not intuitive for understanding network data. Boxplots separated by the many outdegree values observed do not communicate \textit{how} the nodes are connected, just that some have more connections than others. In order to understand the fit of the model, we should try to understand how well the model captures the\textit{overall} structure of the network, not just one or two summary measurements of that structure.} %The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data.
%In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.

@@

% removed the following plot
<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3, eval = FALSE>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = y, y=-x)) +
  geom_tile(aes(fill = as.factor(value))) +
  scale_fill_manual("i likes j", labels=c("no", "yes"),
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymax = -19.5, ymin = -35.5, fill =NA)) +
  facet_wrap(~wave) + theme_bw() +
  labs(x = "node j (alter)", y = "node i (ego)") +
  ThemeNoNet +
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
@

@@

source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))),
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat,
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .05, ecolour = 'black', labelon=T, labelcolour = 'white', fontsize = 2, vjust = .5, linewidth=0.25) +
  scale_color_manual(values = cols[8:5], name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") +
  theme(legend.position = 'bottom',
        panel.background = element_rect(fill = NA, color = "grey30"),
        legend.text = element_text(size = 10, family = "Times New Roman"),
        legend.title = element_text(size = 10, family = "Times New Roman"))
alldataviz
@

@@

set.seed(12345)
ggplot(data = jTT, aes(from_id = from, to_id = to)) +
  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T,
           labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15,
           size=10, fontsize = 5, ecolour="black",
           linetype = c("solid", "solid","dotted", "solid")) +
  expand_limits(x=c(-0.2,1.2), y=c(-0.1,1.1)) +
  ThemeNet +
  theme(legend.position = "none") + 
  labs(title = expression(paste(s[i3], "(x,z)")))
  # labs(title = expression(paste(s[i3], "(x,z)", "=",Sigma[j != h], x[ij], x[ih], x[hj] %*% 
  #                               I, "(", z[i], "=", z[h]!= z[j], ")")
  #                         )
  #      )
@
\caption{\label{fig:jtt}Realization of a JTT, where the covariate is represented by the shape of the nodes. The dotted tie is encouraged if $\beta_3$ is positive. $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$}
\end{subfigure}\hfill
\begin{subfigure}[t]{.45\textwidth}
<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\textwidth'>>=
dade <- data.frame(from = c('i', 'i', 'h', 'k', 'i'), to = c('h', 'k', 'j', 'j','j'))
dadn <- data.frame(id = letters[8:11], group = c(1,1,1,1))

@@

set.seed(12345)
ggplot(data = dad, aes(from_id = from, to_id = to)) +
  geom_net(directed = T, labelon = T, ecolour="black",
           linetype = c("solid", "solid", "dotted", "solid","solid" ,"solid"),
           labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15, size=10, fontsize = 5) +
  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
  ThemeNet +
  theme(legend.position = "none") + 
  labs(title = expression(paste(s[i4], "(x)")))
  # labs(title = expression(paste(s[i4], "(x)", "=", "|{j:", x[ij],"=0, ", Sigma[h], x[ih], x[hj] >= 2, "}|" )
  #                         )
  #      )
@
\caption{\label{fig:dad}Realization of a N22 between actors $i$ and $j$. The dotted tie is encouraged if $\beta_4$ has a positive value. $s_{i4}(x) =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$}
\end{subfigure}
\caption{\label{fig:structures} The additional network effects included in our models fit to the friends data. On the left, a JTT between $i$ and $j$, who have different covariate values, and on the right, a N22 between $i$ and $j$, who are indirectly connected by nodes $k$, $h$.}
\end{figure}

@@

#762a83
#af8dc3
#e7d4e8
#d9f0d3
#7fbf7b
#1b7837

@@

# histogram of number of MS in 1000 reps
# sfmsall %>% group_by(rep) %>% summarize(noMs = max(ms)) %>% ggplot() + geom_histogram(aes(x = noMs), binwidth = 5)

sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")

>>>>

<<setup, echo = FALSE, message = FALSE, warning = FALSE, purl=FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\textwidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, dev="cairo_pdf", fig.width = 6, fig.height = 6)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
@

@@

<<getdata, echo=FALSE>>=
set.seed(823746)
load("data/ansnullpaper.rda")
ansnullchains <- get_chain_info(ansnull)
load("data/M1sims1000.RData")
load("data/M2sims1000.RData")
load("data/M3sims1000.RData")
@

\section{Introduction \& Background}
% peices of old intro
%Social networks, such as collaboration networks between academic researchers or friendship networks 
% new intro 
%As humans and researchers, it is our instint to try to learn as much as we can about the world around us. 
As members of the human family, one of the hardest things we try to do is understand our relationships with one another, whether as friends, family, or colleagues. These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals. There are also statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}. These static models, however, are not optimal for modeling ``real-world" social networks because they do not account for changes in the network as time passes. \textit{Dynamic social networks}, which are observed at many points in time, are the alternative to static networks.

Models for dynamic social networks, however, are more complicated than those for static networks. These models must account for both the structure and the change in time.  But if we can capture the network change mechanisms in time, % are more realistic because they provide mechanisms for analyzing multiple observations of the same network over time.

%., and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms. %Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our relationships with each other.

\par One family of models for dynamic social networks are continuous time Markov chain (CTMC) models, introduced in \citet{ctmcorig}. These models condition on the past to model the present state of the network. %These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree, common in ERGMs.
%These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information. Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example 
With these models, we can examine assumptions made about social networks, i.e. that people with common interests are more likely to become friends. For example, we may assume that teenagers who drink heavily are more likely to be friends with other tennagers who drink heavily. The SAOM family allows us to incorporate this behavioral information on shared interests into the modeling process.

@@

<<toyex, echo = FALSE>>=
set.seed(81393)
nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
@
<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the corresponding adjacency matrix visualization.", out.width='.49\\textwidth', fig.height=4>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) +
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
  ThemeNet +
  theme(plot.title = element_text(size = 20)) +
  labs(title = "Node-Link Diagram")

ggplot(data = toynet$edges) +
  geom_tile(aes(x = to, y = -from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) +
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) +
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  ThemeNoNet +
  theme(panel.grid = element_blank(), aspect.ratio = 1,
        plot.title = element_text(size = 20),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        axis.text.x  = element_text(size = 18),
        axis.text.y  = element_text(size = 18)) +
  labs(title = "Adjacency Matrix Visualization", x = "\nto (alter)", y = "from (ego)\n")
@

@@

In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probabilities $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. These probabilities are defined via the objective function. Let $x(i \leadsto j)$ be the network identical to network $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$; i.e.\ the presence or absence of node $x_{ij}$ is flipped in the new network. The probability that the tie $x_{ij}$ changes is defined as:
\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
\end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
%When the value of the objective function is high at the current state, the probability of not making a change is also high.
We explore these probabilities further in Section~\ref{sec:algorithms}. The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change. For more details on CTMCs, see e.g. \citet{ctmc}.

@@

@@

% removed the following plot
<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3, eval = FALSE, purl=FALSE>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = y, y=-x)) +
  geom_tile(aes(fill = as.factor(value))) +
  scale_fill_manual("i likes j", labels=c("no", "yes"),
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymax = -19.5, ymin = -35.5, fill =NA)) +
  facet_wrap(~wave) + theme_bw() +
  labs(x = "node j (alter)", y = "node i (ego)") +
  ThemeNoNet +
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
@

@@

source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))),
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
alldat$X1 <- as.factor(alldat$X1)
alldat$X2 <- as.factor(alldat$X2)
alldataviz <- ggplot(data = alldat,
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .05, ecolour = 'black', labelon=T, labelcolour = 'white', fontsize = 2, vjust = .5, linewidth=0.25) +
  scale_color_manual(values = cols[8:5], name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") +
  theme(legend.position = 'bottom',
        panel.background = element_rect(fill = NA, color = "grey30"),
        legend.text = element_text(size = 10, family = "Times New Roman"),
        legend.title = element_text(size = 10, family = "Times New Roman"))
alldataviz
@

@@

set.seed(12345)
ggplot(data = jTT, aes(from_id = from, to_id = to)) +
  geom_net(aes(shape = as.factor(group)), directed = T, labelon = T,
           labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15,
           size=10, fontsize = 5, ecolour="black",
           linetype = c("solid", "solid","dotted", "solid")) +
  expand_limits(x=c(-0.2,1.2), y=c(-0.1,1.1)) +
  ThemeNet +
  theme(legend.position = "none") + 
  labs(title = expression(paste(s[i3], "(x,z)")))
@
\caption{\label{fig:jtt}Realization of a JTT, where the covariate is represented by the shape of the nodes. The dotted tie is encouraged if $\beta_3$ is positive. $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$}
\end{subfigure}\hfill
\begin{subfigure}[t]{.45\textwidth}
<<dab, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center', out.width='.49\\textwidth'>>=
dade <- data.frame(from = c('i', 'i', 'h', 'k', 'i'), to = c('h', 'k', 'j', 'j','j'))
dadn <- data.frame(id = letters[8:11], group = c(1,1,1,1))

@@

set.seed(12345)
ggplot(data = dad, aes(from_id = from, to_id = to)) +
  geom_net(directed = T, labelon = T, ecolour="black",
           linetype = c("solid", "solid", "dotted", "solid","solid" ,"solid"),
           labelcolour='white',vjust = 0.5, hjust =0.5, arrowgap = .15, size=10, fontsize = 5) +
  expand_limits(x=c(-0.1,1.1), y=c(-0.1,1.1)) +
  ThemeNet +
  theme(legend.position = "none") + 
  labs(title = expression(paste(s[i4], "(x)")))
@
\caption{\label{fig:dad}Realization of a N22 between actors $i$ and $j$. The dotted tie is encouraged if $\beta_4$ has a positive value. $s_{i4}(x) =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$}
\end{subfigure}
\caption{\label{fig:structures} The additional network effects included in our models fit to the friends data. On the left, a JTT between $i$ and $j$, who have different covariate values, and on the right, a N22 between $i$ and $j$, who are indirectly connected by nodes $k$, $h$.}
\end{figure}

@@

@@

# histogram of number of MS in 1000 reps
sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")
